Evaluating the Performance and Energy Efficiency of the \textsc{COSMO-ART} model system

### 2014-04-25 trunk/paper_enahpc14/Sec3:Environment-setup.tex
Manuel's changes

Benchmarks presented  in this work  were performed on  three different
platforms  and can  be reproduced  within an  expected  variance, thus
allowing a  fair comparison between the baseline  and future milestone
versions of \textsc{Cosmo-art}.

\begin{itemize}
\item \textsc{Monch} is a 10 rack NEC-provided cluster composed
  of 312 compute  nodes and utilized by members  of the Swiss Platform
  for Advanced Scientific  Computing (PASC) project~\citep{PASC} at CSCS. Each
  node comprises two Intel Ivy Bridge EP  E5-2660 ten-core processors
  operating at  2.2 GHz, equipped with 32\,GB of DDR3 1600\,MHz  RAM and connected  via InfiniBand Mellanox  SX6036 and FDR switches. For our  experiments, 
  a full rack constituted  of  52
  standard compute  nodes was used.  \textsc{Monch} is  slated to stay
  in  service  without  hardware  upgrade  for  the  duration  of  the
  Exa2Green  project,  allowing   an  identical  configuration  to  be
  recreated for future assessments of the baseline.\\

\item \textsc{Pilatus} is a dual-socket eight-core Intel Sandy
  Bridge EP based cluster composed of 42 compute nodes and used as Piz
  Daint pre-post  processing cluster at  CSCS.  The 2 login  nodes and
  the 42 computes nodes consists in 11 twin-pair Intel E5-Series DALCO
  r2264i4t 2U scalable compute modules. Each module contains 4 compute
  nodes based  on two Intel Xeon E5-2670 processors operating  at 2.6\,GHz 
  connected  via InfiniBand Mellanox SX6036 and FDR  switches (56
  Gb/s speed), and is equipped with 64GB of DDR3 1600\,MHz RAM. For our
  experiments, a  full rack constituted  of 42 standard  compute nodes
  was  used. \textsc{Pilatus}  is based  on  Intel's second-generation
  processors, conventional in  HPC systems and known to  be more power
  consuming than Ivy Bridge successor.\\

\item \textsc{Tintorrum} is  a heterogeneous cluster composed of
  28 compute nodes at UJI. For our  experiments only a subset of 16 nodes was considered,  
  Each of nodes of this subset includes a dual-socket hexa-core Intel Xeon E5645 processors 
  running at 2.40\,GHz, equipped with 288\,GB  of DDR3 1333\,MHz and connected via Infiniband QDR with a Mellanox MTS3600 switch. A  full tracing  experiment is  conducted on
  \textsc{Tintorrum} in order to  capture an overall  power profile at a much
  finer resolution. Also we analize the contribution of the MPI library
  to the energy consumption and how the use of the blocking versus polling
  message-passing policies can potentially render energy savings.
\end{itemize}

@@

\subsubsection{Power-performance measurement framework}

@@

\textsc{COSMO-ART}  is compiled  using the  Extrae  compiler wrappers,
which automatically  instrument the Fortran code of  the model.  Next,
\textsc{COSMO-ART}  is  run on  the  nodes,  thus dissipating  certain
amount  of power.   These  nodes are  connected  to power  measurement
devices that account for the dissipated power/consumed energy and send
the power  data to the  tracing server.  The client,  meanwhile, sends
start/stop  primitives  in  order  to  gather  captured  data  by  the
watt-meters onto the  tracing server, where an instance  of the \pmlib
server is running.

Once  the run  is finished,  a file  containing the  power  profile is
created using the power data  received from the tracing server and the
instrumentation post-processing generates the performance trace files.
All  these files  are  combined then  in  Paraver which  allows us  to
visualise   the   performance  trace   and   the   power  profile   of
\textsc{Cosmo-art} all together.

@@

The \textsc{Cosmo-art}  baseline is a  pure MPI based  Fortran~90 code
currently  running  on   distributed  multi-core  systems  only.   The
software stack on the platforms was controlled using the modules
framework which gives an easy  and flexible mechanism to access to all
of the  provided compilers, tools  and applications.  For  our initial
benchmarking,   we  opted   for   the  GNU   compiler  gcc-4.8.1   on
\textsc{Monch} and gcc-4.8.2 on  \textsc{Pilatus} and \textsc{Tintorrum}) using the -O3 compiler
flag  in favor  of the  Intel compilers 14.0.1,  delivering inferior
performance.  In  addition, we installed the following implementations of MPI: MVAPICH2 1.9 on \textsc{Monch} and \textsc{Pilatus} and  OpenMPI 1.6.5 on \textsc{Tintorrum}. On the other hand, we  used HDF5 1.8.12 and NetCDF  (4.3.1) libraries for  the management  of extremely  large and
complex data collections.  All computes nodes have an operating system
based on GNU/Linux featuring ``2.6.32-358.11.1.el6.x86\_64'' kernel in
\textsc{Monch}     and      ``3.0.101-0.15-default''     kernel     in
\textsc{Pilatus}.

A snapshot of the code,  which includes at least conceptually, all the
information needed  to reproduce the  energy-to-solution benchmarks of
\textsc{COSMO-ART}, was produced and run  on a 1040 cores using 20~MPI
tasks  per node on  \textsc{Monch} and  on a  1344 cores  using 16~MPI
tasks per node on  \textsc{Pilatus}.  The calculated region was mapped
to the participating processors using a 2D-partitioning strategy.  The
distribution along the $x$ and $y$ coordinates was defined by setting:
$nprocx=40$  and $nprocy=26$  for \textsc{Monch}  and  $nprocx=28$ and
$nprocy=24$  as  $nprocx$  is   usually  kept  bigger  than  $nprocy$.
Besides,  as this  version doesn't  make use  of the  traditional GRIB
library, we  specified $nprocio=0$  for GRIB I/O.   Hyperthreading was
not considered in this study  as it previously revealed that it always
led to higher energy-to-solution.

>>>>

Benchmarks presented  in this work  were performed on  three different
platforms  and can  be reproduced  within an  expected  variance, thus
allowing a  fair comparison between the baseline  and future milestone
versions of \cosmoart.

\begin{itemize}
\item \monch is a 10 rack NEC-provided cluster composed
  of 312 compute  nodes %and utilized by members  of the Swiss Platform
  %for Advanced Scientific  Computing (PASC) project~\citep{PASC}
   at CSCS. Each
  node comprises two Intel Ivy Bridge EP  E5-2660 ten-core processors
  operating at  2.2 GHz, equipped with 32\,GB of DDR3 1600\,MHz  RAM and connected  via InfiniBand Mellanox  SX6036 and FDR switches. For our  experiments, 
  a full rack constituted of 52 nodes was used. \monch is  slated to stay
  in  service  without  hardware  upgrade  for  the  duration  of  the
  Exa2Green  project,  allowing   an  identical  configuration  to  be
  recreated for future assessments of the baseline.\\

\item \pilat is a dual-socket eight-core Intel Sandy
  Bridge EP based cluster composed of 42 compute nodes %and used as Piz Daint pre-post  processing cluster 
  at  CSCS.  The 2 login  nodes and
  the 42 computes nodes consists in 11 twin-pair Intel E5-Series DALCO
  r2264i4t 2U scalable compute modules. Each module contains 4 compute
  nodes based  on two Intel Xeon E5-2670 processors operating  at 2.6\,GHz equipped with 64GB of DDR3 1600\,MHz RAM and connected  via InfiniBand Mellanox SX6036 and FDR switches. 
  For our  experiments, a  full rack constituted  of 42 standard  compute nodes
  was used. \pilat  is based  on  Intel's second-generation
  processors, conventional in  HPC systems and known to  be more power
  consuming than Ivy Bridge successor.\\

\item \tinto is  a heterogeneous cluster composed of
  28 compute nodes at UJI. For our  experiments only a subset of 16 homogeneous nodes was considered. Each of them is composed dual-socket Intel Xeon E5645 hexa-core processors 
  running at 2.40\,GHz, equipped with 288\,GB  of DDR3 1333\,MHz and connected via Infiniband QDR with a Mellanox MTS3600 switch. A  full tracing  experiment is  conducted on
  \tinto in order to  capture an overall  power profile at a much
  finer resolution. In this sense, we also analyze the contribution of the MPI library
  to the energy consumption and how the use of the blocking versus polling
  message-passing policies can potentially render energy savings.
\end{itemize}

@@

\subsubsection{Power-performance tracing framework}

@@

\cosmoart  is compiled  using the  Extrae  compiler wrappers,
which automatically  instrument the Fortran code of  the model.  Next,
\cosmoart  is  run on  the  nodes,  thus dissipating  certain
amount  of power.   These  nodes are connected to power distribution units (PDUs)
APC8653 that account for the dissipated power with a sampling frequency of 1\,Hz. The power data captured of these PDUs is afterwards collected by the tracing server. The client,  meanwhile, sends start/stop  primitives  in  order  to  gather  captured  data  by  the watt-meters onto the  tracing server, where an instance  of the \pmlib server is running.

Once  the run  is finished,  a file  containing the  power  profile is
created using the power data  received from the tracing server and the
instrumentation post-processing generates the performance trace files.
All  these files  are  combined then  in  Paraver which  allows us  to
visualise   the   performance  trace   and   the   power  profile   of
\cosmoart all together.

@@

The \cosmoart  baseline is a  pure MPI based  Fortran~90 code
currently  running  on   distributed  multi-core  systems  only.   The
software stack on the platforms was controlled using the modules
framework which gives an easy  and flexible mechanism to access to all
of the  provided compilers, tools  and applications.  For  our initial
benchmarking,   we  opted   for   the  GNU   compiler  gcc-4.8.1   on
\monch and gcc-4.8.2 on  \pilat and \tinto) using the -O3 compiler
flag  in favor  of the  Intel compilers 14.0.1,  delivering inferior
performance.  In  addition, we installed the following implementations of MPI: MVAPICH2 1.9 on \monch and \pilat and  OpenMPI 1.6.5 on \tinto. On the other hand, we  used HDF5 1.8.12 and NetCDF  (4.3.1) libraries for  the management  of extremely  large and
complex data collections.  All computes nodes have an operating system
based on GNU/Linux featuring ``2.6.32-358.11.1.el6.x86\_64'' kernel in
\monch, ``3.0.101-0.15-default'' in \pilat and ``2.6.18-238.19.1.el5.x86\_64'' in \tinto.

A snapshot of the code,  which includes at least conceptually, all the
information needed  to reproduce the  energy-to-solution benchmarks of
\cosmoart, was produced and run  on a 1040 cores using 20~MPI
tasks  per node on  \monch and  on a  1344 cores  using 16~MPI
tasks per node on  \pilat.  The calculated region was mapped
to the participating processors using a 2D-partitioning strategy.  The
distribution along the $x$ and $y$ coordinates was defined by setting:
$nprocx=40$  and $nprocy=26$  for \monch  and  $nprocx=28$ and
$nprocy=24$  as  $nprocx$  is   usually  kept  bigger  than  $nprocy$.
Besides,  as this  version doesn't  make use  of the  traditional GRIB
library, we  specified $nprocio=0$  for GRIB I/O.   Hyperthreading was
not considered in this study  as it previously revealed that it always
led to higher energy-to-solution.
